---
title: "Penguin Measurements Across R and Python"
author: "Anushka Seth"
date: 2026-09-24
format:
  html:
    toc: true
engine: knitr
execute:
  echo: true
  warning: false
  message: false
---

## Introduction

This analysis uses the [Palmer Penguins dataset](https://allisonhorst.github.io/palmerpenguins/)
from Palmer Station Antarctica LTER.

We investigate two questions:

1. Does the relationship between bill length and bill depth change when
   penguins are grouped by species?
2. How does average body mass relative to flipper length differ across
   species and sex?

R loads the dataset and calculates the overall correlation. Python accesses
the R data object, calculates species-level correlations, and creates a
mass-to-flipper-length ratio. The resulting data objects return to R for
ranking and visualization.

## Setup

The document uses the `knitr` engine. Python chunks run through
`reticulate`, allowing objects to be shared between languages.

```{r}
#| label: setup

library(reticulate)
library(palmerpenguins)
library(dplyr)
library(ggplot2)

# Use a plain data frame for straightforward conversion to pandas.
penguins_data <- as.data.frame(palmerpenguins::penguins)
```

## Calculation 1: Overall correlation in R

We use observations with both bill measurements available.

```{r}
#| label: overall-correlation-r

overall_corr <- cor(
  penguins_data$bill_length_mm,
  penguins_data$bill_depth_mm,
  use = "complete.obs"
)

overall_results <- data.frame(
  complete_pairs = sum(
    complete.cases(
      penguins_data[, c("bill_length_mm", "bill_depth_mm")]
    )
  ),
  correlation = round(overall_corr, 3)
)

knitr::kable(overall_results)
```

## Calculation 2: Species-level correlations in Python

The expression `r.penguins_data` accesses the data frame created in R.
`reticulate` converts it to a pandas DataFrame.

We also access the R scalar `overall_corr` to compare the pooled correlation
with the species-level results.

```{python}
#| label: species-correlations-python

import pandas as pd

# Transfer the R data frame into Python.
penguins_py = r.penguins_data.copy()

correlation_rows = []

for species, group in penguins_py.groupby("species", observed=True):
    complete = group.dropna(
        subset=["bill_length_mm", "bill_depth_mm"]
    )

    correlation_rows.append({
        "species": str(species),
        "complete_pairs": len(complete),
        "correlation": complete["bill_length_mm"].corr(
            complete["bill_depth_mm"]
        )
    })

species_correlations = pd.DataFrame(correlation_rows)

print(f"Overall correlation from R: {float(r.overall_corr):.3f}")
print(species_correlations.round(3).to_string(index=False))
```

The pooled correlation is negative, but each species has a positive
correlation. This reversal illustrates Simpson's paradox: combining
groups can produce a relationship that differs from the relationships
within those groups.

Species differ in their typical bill dimensions, so the pooled association
mixes within-species and between-species variation.

### Visualizing the reversal in R

Colored lines show separate linear fits for each species. The dashed
black line shows the pooled fit across all species.

```{r}
#| label: fig-simpsons-paradox
#| fig-cap: "Bill measurements show a negative pooled association but positive associations within species."

bill_data <- penguins_data %>%
  filter(
    !is.na(bill_length_mm),
    !is.na(bill_depth_mm)
  )

ggplot(
  bill_data,
  aes(
    x = bill_length_mm,
    y = bill_depth_mm,
    colour = species
  )
) +
  geom_point(alpha = 0.65) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_smooth(
    aes(group = 1),
    method = "lm",
    se = FALSE,
    colour = "black",
    linetype = "dashed"
  ) +
  labs(
    title = "Simpson's Paradox in Penguin Bill Measurements",
    subtitle = "Colored lines: species fits; dashed black line: pooled fit",
    x = "Bill length (mm)",
    y = "Bill depth (mm)",
    colour = "Species"
  ) +
  theme_minimal()
```

## Calculation 3: Mass-to-flipper-length ratio in Python

For consistency with the original analysis, we call this ratio BCI:

$$
\mathrm{BCI} =
\frac{\text{body mass (g)}}{\text{flipper length (mm)}}.
$$

This ratio has units of grams per millimeter. It is a descriptive measure,
not a validated health indicator.

We calculate the ratio for individual penguins, then average it within
species–sex groups. Records with missing sex or missing BCI are excluded
from these group summaries.

```{python}
#| label: bci-calculations-python

penguins_bci = penguins_py.copy()

penguins_bci["bci"] = (
    penguins_bci["body_mass_g"]
    / penguins_bci["flipper_length_mm"]
)

bci_complete = penguins_bci.dropna(
    subset=["species", "sex", "bci"]
)

bci_profile = (
    bci_complete
    .groupby(["species", "sex"], observed=True)
    .agg(
        mean_bci=("bci", "mean"),
        sd_bci=("bci", "std"),
        n=("bci", "count")
    )
    .reset_index()
)

print(bci_profile.round(3).to_string(index=False))
```

## Calculation 4: Bring Python results back into R and rank groups

The `py$` interface accesses objects created in Python. Here we retrieve
both the enriched individual-level dataset and the group summary.

Groups with equal average BCI receive the same rank.

```{r}
#| label: rank-bci-r

# Transfer Python data frames back into R.
penguins_enriched <- py$penguins_bci
bci_summary_r <- py$bci_profile

bci_ranked <- bci_summary_r %>%
  mutate(rank = min_rank(desc(mean_bci))) %>%
  arrange(rank, species, sex) %>%
  select(rank, species, sex, n, mean_bci, sd_bci)

knitr::kable(
  bci_ranked,
  digits = 3,
  caption = "Species–sex groups ranked by average BCI"
)

# Verify that the Python-created column returned successfully.
stopifnot(
  "bci" %in% names(penguins_enriched),
  nrow(penguins_enriched) == nrow(penguins_data)
)
```

### Visualizing Python results in R

```{r}
#| label: fig-bci-comparison
#| fig-cap: "Mean mass-to-flipper-length ratio by species and sex, calculated in Python and plotted in R."

ggplot(
  bci_summary_r,
  aes(
    x = species,
    y = mean_bci,
    fill = sex
  )
) +
  geom_col(
    position = position_dodge(width = 0.8),
    width = 0.7
  ) +
  labs(
    title = "Average Mass-to-Flipper-Length Ratio",
    subtitle = "Only records with known sex and nonmissing BCI are included",
    x = "Species",
    y = "Mean BCI (g/mm)",
    fill = "Sex"
  ) +
  theme_minimal()
```

## Discussion

The bill measurements demonstrate Simpson's paradox: the overall negative
correlation reverses direction when the data are analyzed within species.
The pooled relationship therefore does not describe the relationship
within any individual species.

Average BCI also differs across species–sex groups. However, a larger
ratio should not automatically be interpreted as better condition.
Species and sexes differ naturally in body size and proportions, and
dividing mass by flipper length does not necessarily remove those
differences.

This workflow demonstrates two-way data exchange:

- **R → Python:** `penguins_data` and `overall_corr`.
- **Python → R:** `penguins_bci` and `bci_profile`.

R handles loading, ranking, and visualization, while Python performs
species-level correlations and BCI calculations.